---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Introduction 

This document contains the first assignment of the Coursera Reproducible Research course, dated 07/01/2018. The aim is to read in a data set, preprocess it to a usage stage and present an analysis and further extension of the data set.

This section loads the appropriate libraries
```{r}
library(ggplot2, readr);library(dplyr);library(tidyr)
```

## Loading and preprocessing the data

The existing dataset has variables: 
*steps

*date

*interval (representing the 5-minute interval the steps are counted in). The data is read in as is. Specific data tables are made to use for the total step count analysis and the step-count by time analysis

```{r, echo=TRUE}
datStore = read.csv("F:/RepData_PeerAssessment1/activity/activity.csv")
totalStepStore <-aggregate(data=datStore,steps~date,sum)
colnames(totalStepStore) <- c("date", "StepsOnDay")
intervalAvg <- aggregate(data=datStore,steps~interval,median)
colnames(intervalAvg) <- c("interval", "steps")
```

## What is mean total number of steps taken per day?

For this section we want a histogram of the total number of steps per day, and the mean and median number of steps per day. Firstly a histrogram of total number of steps per day.

```{r, echo=TRUE}
ggplot(data=totalStepStore, aes(totalStepStore$StepsOnDay))+geom_histogram()
```
And the mean and median number of total steps per day is
```{r, echo=TRUE}
mean(totalStepStore$StepsOnDay)
median(totalStepStore$StepsOnDay)
```

## What is the average daily activity pattern?

```{r, echo=TRUE}
ggplot(data=intervalAvg, aes(interval, steps))+geom_line()+ xlab("# Interval")+ylab("Steps in 5 minute interval")
```

The interval number with the highest average steps is 

```{r, echo=TRUE}
intervalAvg$interval[which(intervalAvg$step==max(intervalAvg$steps, na.rm=TRUE))]
```

## Imputing missing values

There's a number of NA values in the dataset. The summary of datStore$steps is :
```{r, echo=TRUE}
summary(datStore$steps)
```
and we can see there are 2304 NA values. To fill this in we will use the median number of steps for that interval value. A subfunction that will check is steps is NA in full table, and if so replace with median, will be used. 

```{r, echo=TRUE}
filledDatStore <-datStore
replace_unfilled = function(tarOne, tarTwo){
  if(is.na(tarOne)){
    return(tarTwo)
  }
  return(tarOne)
}
filledDatStore$steps = mapply(replace_unfilled, datStore$steps, intervalAvg$steps)
summary(filledDatStore)
```
The summary shows that there are now no NA values \o/.

The updated histogram of information is

```{r, echo=TRUE}
intervalAvgFilled <- aggregate(data=filledDatStore,steps~interval,median)
colnames(intervalAvgFilled) <- c("interval", "steps")

ggplot(data=intervalAvgFilled, aes(interval, steps))+geom_line()+ xlab("# Interval")+ylab("Steps in 5 minute interval")

```

And the mean and median values are
```{r, echo=TRUE}
totalStepStore <-aggregate(data=filledDatStore,steps~date,sum)
mean(totalStepStore$steps)
median(totalStepStore$steps)
```

As a consequence of this, the mean total steps per day drops from 10766 (excluding NAs) to 9503 (more than 10%, a small decrease) and median from 10765 to 10395, a less significant decrease.

## Are there differences in activity patterns between weekdays and weekends?

First add a factor for weekday or weekend

```{r, echo=TRUE}
filledDatStore$dayname <- weekdays(as.Date(filledDatStore$date))
head(filledDatStore)

typeOfDay = function(dayname){
  if(dayname %in% c("Saturday","Sunday")) {
    return("weekend")
  }
  return("weekday")
}

filledDatStore$daytype = as.factor(apply(as.matrix(filledDatStore),1,typeOfDay))

```
Now plot weekday time series and weekend time series to show differences in activity. New data frames for each type will be created.

```{r, echo=TRUE}
ggplot(data=filledDatStore, aes(interval, steps,group=daytype))+geom_line()
```